---
title: 'R Notebook: Multiple Regression Model of Student Academic Achievement'
author: "Tristen Bristow"
date: "April 6, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###  Abstract  
#### The interest of this study is in developing a prediction model of student success  
#### based on measured factors of success in Mathematics and Portuguese. Multiple  
#### regression is applied to develop a regression classifier based on student-provided  
#### factors that relate to living conditions and education conditions. Though regression  
#### tree modeling would appear at first to be the correct approach, the number and  
#### high-cardinality nature of many of the variables in this data makes such an approach  
#### less feasible in practice.  



```{r one, message=FALSE, warning=FALSE}
set.seed(1)
library(car)
library(boot)
c <- read.table("student-por.csv",sep=";",header=TRUE)
c <- data.frame(c)
d <- read.table("student-mat.csv",sep=";",header=TRUE)
d <- data.frame(d)
e <- rbind(c,d)
```

### Data Cleaning  

#### Both Math and Porteguese sets are merged, alternate column titles are applied,  
#### and all student grades are averaged across three grade entries.  

```{r two, message=FALSE, warning=FALSE}
names(e) <- c("school","sex","age", "address","family size","parents cohab.", "mom's education", 
             "dad's education","mom's job", "dad's job","reason", "guardian","travel", "study", 
             "failures","education support","family support","paid","activities", "nursery","higher",    
             "internet","romantic","family bond","free time","social","workday alch.","weekend alch.","health",
             "absences","Grade 1","Grade 2","Grade 3")   

Grade <- (e$`Grade 1` + e$`Grade 2` + e$`Grade 3`) / 3
e$`Grade 1` <- NULL 
e$`Grade 2` <- NULL 
e$`Grade 3` <- NULL
e = cbind(e, Grade)
attach(e)
```

###  Exploratory Investigation  
#### From initial inspection it is clear education success is quantified by the Grade  
#### variable. Results of inspection indicate general normality of this output variable.  
#### The class distributions of explanatory variables, 'dad's job' and 'mom's job' appear 
#### to show questionable value by inspection of the summary table. This is indicated   
#### by the limited difference between class-levels, except for the vaguely defined class, 
#### 'other', showing the survey question isn't well-defined or reliable an indicator. 
#### An inspection of the VIF's (Variance Inflation Factors), of model parameters is 
#### performed to check for multicolonarity in the dataset.  
 
```{r three}
summary(e)
hist(Grade)
str(e)
names(e)

x1 <- e[c(31, 1 : 10)]
pairs(x1)
x2 <- e[c(31, 10 : 20)]
pairs(x2)
x3 <- e[c(31, 20 : 30)]
pairs(x3)
```

###  Model Development  
#### K-fold CV is applied in the fitting of linear models to the training data. Successively,  
#### models of lesser complexity are derived (starting with the saturated model), selecting  
#### statistically significant predictors that are reported with every model fit. Cross  
#### validation indicates an Mean Square Error rate estimate to verify that, in choosing  
#### lower complexity models, we are not introducing significant error. Finally, a best-  
#### fit model containing significant predictors (showing little difference in MSE from  
#### the saturated model), is tested with the clean data to provide an out-of-sample  
#### estimate for model performance.  
  
#### The following is a series of progressive model fits performed find the best possible   
#### fit. The saturated model cardinality is 30 variables. 10-fold CV is applied to the MSE
#### estimation of model performance on the test data.  
```{r five}
names(e)
fit <- glm(Grade~., data = e)
MSE1 <- cv.glm(e, fit, K = 10)$delta[1]
summary(fit)
```
#### A report on statistical significance of saturated model coefficients indicates  
#### significant (p < 0.01) predictors of Grade to be study, failures, education, 
#### support, paid, and higher. 
### MSE for 10-Fold CV of fit of saturated model:  
```{r fivehalf}
MSE1
```
#### A check for multicolinarity by VIF shows negative results, indicating the potential for  
#### linear modelling success (conditioned on all GVIF values being less than 10).  

```{r six}
vif(fit)
```
 
### Second Fit:  
#### Now a lower complexity model of 5 variables (reported significant), from the saturated  
#### model is fitted. All variables included are checked for significance (p < 0.01).  

```{r seven}
fit2 <- glm(Grade ~ study + failures + `education support` + paid + higher, data = e)
MSE2 <- cv.glm(e, fit2, K = 10)$delta[1]
summary(fit2)
```

### 10-Fold CV MSE estimate of Model II:  

```{r sevenhalf}
MSE2
```

#### This model shows negligible difference in MSE from the saturated model and contains  
#### 26 fewer predictors, thus indicating potential for out-of-sample performance.  

```{r eight}

summary(fit2)
par(mfrow = c(3, 2))

plot(Grade, study)
plot(Grade, failures)
plot(Grade,`education support`, yaxt='n')
axis(2, labels = c("false","true"), at = c(1, 2)) 
plot(Grade, paid, yaxt='n')
axis(2, labels = c("false","true"), at = c(1, 2)) 
plot(Grade, higher, yaxt='n')
axis(2, labels = c("false","true"), at = c(1, 2)) 
mtext("Significant Factor Plots", side = 3, line = -3, outer = TRUE)


```

###  Results  
#### This results of this work should be seen as a starting point for more advanced studies  
#### of success prediction in general education. They may only hold significance for the  
#### originating educational department. It appears strictly domain specific, general  
#### claims to any general predictive success of any derived models is not generally indicated.  
#### This view is surmised from the collected data and available documentation. The intent  
#### interpreted is to find a model of specific factors relevant to learning success that are  
#### shared between Mathematics an Portuguese, discussed in identical terms - identical variables  
#### are chosen for both data sets as collected from student surveys. This indicates an  
#### implicit assumption of the study: a uniform learning measure exists between mathematics  
#### and language. This assumption appears to be latent in the study variables, as chosen factors  
#### are more generally living-condition or non-subject specific. It is apparent that for the  
#### chosen causal-factors the data is not sufficiently diverse in academic types to address  
#### the generality of their scope. In addition, this study has a limited data quantity relative  
#### to the number of the predictors. All inclusions/exclusions suggested by this regression  
#### model should be seen as restricted from any claims of predictive power in a broader  
#### or differing range or academic subjects.  
####  
###  Conclusion  
#### This work could be seen a good use of resources in determining how best to design future  
#### studies, specifically, what questions to exclude from study surveys in any following work.  
#### Results indicate a reduction in the set of explanatory variables by a factor of 5 in  
#### predictive modeling. Certain difficulties are present in interpreting results from the  
#### data documentation, indicating restricted application to the listed subjects.  


